Temporality is Important for Networks of Hospital Patient Transfers

### 2024-11-12 reports/report.qmd
Update report

---
title: "Temporality is Important for Networks of Hospital Patient Transfers"
authors:
    - name: David Wu
    - name: Michael Lydeamore
    - name: Andrew Stewardson
    - name: Tjibbe Donker
    - name: Nic Geard
    - name: "******"
date: last-modified
format:
    html: default
    pdf:
        include-in-header:
            - text: |
                \usepackage{mathtools}
        papersize: a4
bibliography: refs.bib
csl: plos.csl
jupyter: python3
# filters:
#     - diagram
execute:
    echo: false
    warning: false
    error: false
---

```{python}
import numpy as np
import polars as pl
from matplotlib import pyplot as plt
import seaborn as sns
import itertools
import re

@@

# Glossary

@@

define a vertex $v_i = (x_i, t_i)$ made up of a location and time resp.

@@

<- yank from [@lydeamore_carbapenemase-producing_2024]

Admissions data is from the Victorian Admitted Episdoes Dataset (VAED).

Some exploratory analysis was done in by [@lydeamore_carbapenemase-producing_2024]

@@

Patient transfers can be considered as edges from some source hospital $x_i$ at some time $t_i$ to a target hospital $x_j$ at some time $t_j$. We could work with the exact times, but this will lead to an overly sparse representation that is difficult to work with.
Instead, we discretise time into bins by treating movements involving some arbitrary hospital $x_i$ between times $t_i$ and $t_i + W$ as instead involving a single node at $v_i \coloneqq (x_i, t_i)$. 
Then we can have a temporal network that can be treated as a static network as in @sec-network-modelling, i.e. we have edge weights $w_{ij}$ along the edge $e_{ij} \coloneqq (v_i, v_j)$.

@@

### Dynamics

<!-- TODO need to work on indexing consistency -->

@@

## A Reduced Model

@@

```{dot}
//| fig-width: 50%
digraph G {
    S₀ -> S₁ [ label="λ₀" ]
}
```

@@

In the model without indirect transfer, we can compute the expected hitting time $k$ from $S_0$ to $S_1$

$$k = {\lambda_0}^{-1}$$

In a model with indirect transfer, we introduce an additional state $S_2$ which represents the indirect pathway. If the probability of direct transfer is $p$ given that an individual will move out of $S_0$, and the average indirect transfer period is $T = \lambda_2^{-1}$, the we can compute the hitting time $k$ of $S_1$ given we start at $S_0$:

$$k = {\lambda_0}^{-1} + \frac{1-p}{\lambda_2}$$

@@

For a larger network, the number of possible paths causes complications for explicit computation of the effect, but we can always expect it to follow a similar principle, where splitting transfers into a direct path and a slower indirect path will add to the expected hitting time. These can be obtained by solving a linear system @Norris_Markov_Chains_1997.
For the naïve static network, we can model the rate of movement from hospital $i$ to hospital $j$ as the mean number of observed movements over the observation period (XXX days).
For the improved static network, we expand the state by including an additional state for each pair of hospitals, so that the residence time at home (between hospital admissions) for indirect transfers can be modelled. Then as with the naïve model, the movement from $i$ to $j$ directly is the mean observed number of transfers over the observation epriod. The indirect movements from $i$ to the intermediate $i,j$ state are derived from the observed number of movements, and the rate from $i,j$ to $j$ is derived from the mean readmission period (as seed in @fig-readmission-window).

Comparing the relative difference in the mean hitting times gives us @fig-hitting-time-difference

@@

# temp hack
# todo port across computation code
from PIL import Image
image = Image.open("./precompiled/zrelative_change_hitting_time_home.png")
image
```

@@

>>>>

---
title: "Temporality is Important for Networks of Hospital Patient Transfers"
authors:
    - name: David Wu
    - name: Michael Lydeamore
    - name: Andrew Stewardson
    - name: Tjibbe Donker
    - name: Nic Geard
    - name: "******"
date: last-modified
format:
    html: default
    pdf:
        include-in-header:
            - text: |
                \usepackage{mathtools}
        papersize: a4
bibliography: refs.bib
csl: plos.csl
jupyter: python3
execute:
    echo: false
    warning: false
    error: false
---

```{python}
import numpy as np
import polars as pl
from matplotlib import pyplot as plt, colors
import seaborn as sns
import h5py
import itertools
import re

@@

<!-- # Glossary

@@

define a vertex $v_i = (x_i, t_i)$ made up of a location and time resp. -->

@@

Victoria, Australia, has a population of 7.0 million people [@ABS_state_populations]. Its hospital system consists of more than 300 hospital campuses, which vary in size, case mix and services.

Hospital admissions data was sourced from the Victorian Admitted Episdoes Dataset (VAED) [@VAED_official].
In particular, admissions data from [] to [] ([] days) was used in this study. All Victorian hospital admissions during this study period are included. The total number of admissions is []
This data was aggregated into time-discretised temporal networks (described below), and exported from the database; further analysis as presented in this paper is performed on external machines with no access to identifiable protected health information.

Some exploratory analysis of this data was done in by [@lydeamore_carbapenemase-producing_2024].

@@

Patient transfers can be considered as edges from some source hospital $x_i$ at some time $t_i$ to a target hospital $x_j$ at some time $t_j$. 
Because we cannot export exact times, we instead discretise time.
This is done by treating movements involving some arbitrary hospital $x_i$ between times $t_i$ and $t_i + W$ as instead involving a single node at $v_i \coloneqq (x_i, t_i)$. 
Then we can have a temporal network that can be treated as a static network as in @sec-network-modelling, i.e. we have edge weights $w_{ij}$ along the edge $e_{ij} \coloneqq (v_i, v_j)$, but with some explicit temporal representation.

@@

## Dynamics

@@

### A Reduced Model

@@

```{dot}
//| fig-width: 30%
digraph G {
    S₀ -> S₁ [ label="λ₀" ]
}
```

@@

In the model without indirect transfer, we can compute the expected hitting time $k_{1}$ from $S_0$ to $S_1$ as

$$k_{1} = {\lambda_0}^{-1}$$

In a model with indirect transfer, we introduce an additional state $S_2$ which represents the indirect pathway. If the probability of direct transfer is $p$ given that an individual will move out of $S_0$, and the average indirect transfer period is $T = \lambda_2^{-1}$, the we can compute the hitting time $k_{1^*}$ of $S_1$ given we start at $S_0$:

$$k_{1^*} = {\lambda_0}^{-1} + \frac{1-p}{\lambda_2}$$

@@

For a larger network, the number of possible paths causes complications for explicit computation of the effect, but we can always expect it to follow a similar principle, where splitting transfers into a direct path and a slower indirect path will add to the expected hitting time. These can be obtained by solving a linear system [@Norris_Markov_Chains_1997].

## Temporal Network Analysis

## Effect of Home-return Mechanic

For the naïve static network, we can model the rate of movement from hospital $i$ to hospital $j$ as the mean number of observed movements over the observation period (XXX days).
For the improved static network, we expand the state by including an additional state for each pair of hospitals, so that the residence time at home (between hospital admissions) for indirect transfers can be modelled. Then as with the naïve model, the movement from $i$ to $j$ directly is the mean observed number of transfers over the observation epriod. The indirect movements from $i$ to the intermediate $i,j$ state are derived from the observed number of movements, and the rate from $i,j$ to $j$ is derived from the mean readmission period (as seed in @fig-readmission-window).

Comparing the (uniform) relative difference in the mean hitting times ($= k_{1^*} / (k_1 + 1)$) gives us @fig-hitting-time-difference. We see that there is significant banding in the structure of the matrix.

@@

with h5py.File('../hitting_time_analysis/hitting_time_arrays.h5', 'r') as fp:
    naive_static_times = fp['naive_static'][:]
    delayed_static_times = fp['delayed_static'][:]

relative_change = delayed_static_times / (naive_static_times + 1)
plt.matshow(relative_change, cmap='viridis', fignum=99)
plt.colorbar(label='Relative difference in hitting time')
plt.xlabel('Hospital to hit')
plt.ylabel('Starting hospital');
```

@@

# Discussion

The modelling of patient transfers between hospitals has historically been done by using static networks.
